{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming

### 2025-04-04 paper.tex
Evaluation, conclusions, editing

\begin{abstract}
User-centric programming research gave rise to a variety of compelling programming experiences,
including local-first collaborative editing, programming by demonstration, incremental
recomputation, schema change control, end-user debugging and concrete programming.
Those experiences advance the state of the art of end-user programming, but they are hard to
implement on the basis of established programming languages and system.

We contribute Denicek, a computational substrate that simplifies the implementation of the
above programming experiences. Denicek represents a program as a series of edits
that construct and transform a document consisting of data and formulas. It provides three
operations on edit histories: edit application, merging of histories and conflict resolution.
By composing these operations, we can  easily implement a range of programming experiences.

@@

A computational substrate defines the structures with which programs are constructed, how
the progam state is represented and how it evolves \cite{jakubovic-2022-ladder}. The choice of
a substrate affects what programming experiences can be readily supported by a system. For example,
object-oriented programming has been historically linked to graphical user
interfaces~\cite{kay-1993-smalltalk}, while representing programs as lists enabled Lisp to become
a language laboratory \cite{steele-1993-lisp}.

@@

In this paper, we describe a programming substrate that makes it easy to develop diverse
programming experiences \cite{myers-2006-eup}. Those include:

\begin{itemize}
\item \emph{Local-First Collaboration.} Multiple users should be able to concurrently
  modify a single document and automatically merge their changes, preferrably without a central server~\cite{kleppmann-2019-local,litt-2022-peritext}.
\item \emph{Programming by Demonstration.} Allow users to construct simple programs by
  enacting the steps of the expected behavior using concrete examples~\cite{leiva-2021-rapido,cypher-1993-pbd}.
\item \emph{Incremental Recomputation.} When a part of a document changes, formulas whose result
  depends on it are invalidated and, possibly, automatically recomputed \cite{teitelbaum-1981-cps,mcdirmid-2013-usable,horowitz-2023-engraft}.
\item \emph{Schema Change Control.} When the user evolves the structure of the document, affected
  data and formulas should automatically co-evolve to match the new structure~\cite{litt-2020-cambria,edwards-2025-schema}.
\item \emph{End-User Debugging.} The user should be able to ask provenance questions~\cite{cheney-2009-provenance} --
  why a computation resulted in a particular value and what inputs contributed to the result \cite{ko-2009-whyline}.
\item \emph{Concrete Programming.} It should be possible to reuse parts of program logic, or formulas,
  without introducing abstractions, that is, program against concrete values \cite{edwards-2006-copypaste,edwards-2022-copypaste}.
\end{itemize}

\subsubsection*{Two-Stage Methodology}
The technical focus of this paper fits within the interior mode of design science research \cite{adam-2021-dsr}.
To \emph{design} Denicek, we identify six \emph{formative examples} -- simple programming
tasks that manifest one or more of the desired programming experiences~(\S\ref{app:examples}).
Using those examples, we co-design the Denicek substrate and Webnicek~(\S\ref{sec:walk}),
a simple web-based end-user programming environment built directly on top of the substrate. Although Webnicek can
be used to complete end-user programming tasks, it is optimized for developing the underlying
substrate rather than for usability.

To \emph{evaluate} Denicek, we use the substrate to develop Datnicek (\S\ref{sec:case}), an interactive data
science notebook system inspired by existing environments~\cite{kandel-2011-wrangler,drossos-2020-wrex}.
We use the second stage to evaluate suitability of the Denicek substrate for the development of
end-user programming systems and report the results of our evaluation (\S\ref{sec:eval}).

@@

The key insight that we present in this paper is that many compelling programming experiences can
be implemented by leveraging the uniform document representation alongside with a suitable
composition of the three primitive operations.

Changing data and formulas is done using the same primitive edit operations that
manipulate the document structure. A user-interface may provide a specialized editor, but still
trigger the primitive edits behind the scenes. However, interacting with elements in the document,
such as a entering text in a textbox can also generate a document edit that can be merged or checked
for conflicts (\S\ref{sec:impl-interaction}).

As we will see, past edits that demonstrate an operation done to the
document can be recorded, allowing programming by demonstration (\S\ref{sec:impl-pbd}).  Replaying
such recorded edits is implemented using the merging operation, which means that recorded operations
continue working even if the document structure later evolves. Moreover, structural changes to the
document can be merged with concurrent data edits  (\S\ref{sec:impl-collab}). Evaluation of formulas
also generates document edits (\S\ref{sec:impl-eval}). If the evaluated edits conflict with manual
edits done later by the user, the evaluated edits are removed, implementing an incremental
recomputation (\S\ref{sec:impl-incremental}).

@@

\begin{itemize}
\item We present the Denicek substrate (\S\ref{sec:system}) and provide a detailed description of its
  document representation, edit operations and the key three operations for working with edit histories.
\item We illustrate a range of end-user programming experiences supported in Webnicek, a simple
  web-based prototype programming system (\S\ref{sec:walk}), and discuss how the experiences are
  implemented using the Denicek substrate (\S\ref{sec:impl}).
\item We document important design decisions, alternatives and limitations (\S\ref{sec:discuss}).
  The analysis shows that the desired functionality requires a careful choice among interconnected
  design options.
\item To evaluate how Denicek simplifies the development of program\-ming systems, we
  build an innovative data science notebook Datnicek (\S\ref{sec:case})
  and assess its impementation complexity~(\S\ref{sec:eval}).
\end{itemize}

\noindent
To enable others build on top of Denicek, we share our compact and documented
source code at: \url{https://github.com/removed/for/review} %d3sprog/denicek

@@

\section{Background}
\label{sec:background}
Denicek aims to support a class of systems associated with end-user programming \cite{myers-2006-eup},
notational freedom and self-sustainability~\cite{jakubovic-2023-techdims}, liveness, interactivity
and richness \cite{horowitz-2023-qualities,rein-2019-live}. We see programming more as interacting with a medium
or a substrate \cite{kay-1977-media,klokmose-2015-webstrates} than writing text \cite{rpg-2012-revolution}.
We focus on systems that allow gradual progression from a user to a developer, as envisioned
in Smalltalk~\cite{reenskaug-1981-byte} and use the term end-user programming loosely to refer to a
part of this spectrum, also including spreadsheet systems and notebooks for data science.
We first review related systems and their technical design (\S\ref{sec:background-sys})
before discussing programming experiences (\S\ref{sec:background-exp}).

@@

\subsubsection*{Programming Systems and Substrates}
A number of systems illustrate the qualities Denicek aims to support and have a related internal
structure. Subtext, BootstrapLab and Infra~\cite{edwards-2005-subtext,jakubovic-2022-ladder,hall-2017-infra}
use structured document-based program representation and provide some of the desired programming
experiences on top of this representation. Many of those design ideas can be traced back to
Boxer~\cite{disessa-1986-boxer}, which introduced the \emph{naive realism} principle (what the user
sees is all there is) that we follow in Webnicek.

To indicate that Denicek is intended as an underlying infrastructure on top of which programming
systems can be built, we use the term \emph{computational substrate}, which also dates back to
Boxer~\cite{disessa-1995-epistemology} and is closely related to the notion of dynamic media
introduced by Kay and Goldberg \cite{kay-1977-media}. Webstrates \cite{klokmose-2015-webstrates}
is a substrate based on synchronization of documents (but without edit histories) that has been
use as the basis for multiple programming systems \cite{radle-2017-codestrates,borowski-2022-varv}.

\subsubsection*{Edit Histories and Merging}
Manipulating programs through semantically meaningful edits is a technique used by structure
editors \cite{teitelbaum-1981-cps,hempel-2018-deuce,beckman-2023-sandblocks}. The langauge of
edits has been captured formally as an \emph{edit calculus} \cite{omar-2017-hazelnut} and
edits have also been used as the basis of live programming environments \cite{storm-2013-deltas}.

Merging of edits is most frequently done in version control systems such as git. Systems based
on a more rigorous design such as Pijul \cite{meunier-2024-pijul} delay merging to a later point
by using a graph or a lattice \cite{schurmann-2022-merging}. In the context of
programming environments, Grove \cite{adams-2025-grove} uses a commutative patches, with a graph
structure similar to Pijul, as the basis for a collaborative structure editor.

More generally, merging of edits can be based on the operational transform (OT) approach
\cite{davis-2002-otsgml}, where edit conflicts are reconciled,
or conflict-free representation (CRDTs) \cite{shapiro-2011-crdts,kleppmann-2019-local,litt-2022-peritext}.
In both approaches, supporting complex edits on tree structures
remains a challenge \cite{jungnickel-2016-otjson,da-2024-jsoncrdt}. Mergeable replicated data
types (MRDTs) \cite{kaki-2019-mrdts} can merge updates provided there is a suitable relational
representation of the data. Recent work on MySubstrates and Grove \cite{adams-2025-grove,klokmose-2024-mywebstrates}
has been based on CRDTs, whereas Edwards et al. \cite{edwards-2021-typed} use OT.

@@

\subsubsection*{Local-First Collaboration}
Since the early collaborative programming environments such as Collabode \cite{goldman-2011-collaborative},
real-time collaboration has became widely used, if not always without challenges \cite{tan-2024-vslive}.
Merging concurrent edits is one such challenge.
In addition to CRDT-based approaches \cite{klokmose-2024-mywebstrates}, conflicts arising during
collaboration have been solved through fine-grained locking \cite{wang-2024-nbconflicts}.

\subsubsection*{Programming by Demonstration}
Earliest PbD systems used the paradigm for tasks ranging from graphics
and user interfaces to general-purpose programming \cite{smith-1975-pygmalion,cypher-1993-pbd}.
Wrangler \cite{kandel-2011-wrangler} showed the effectivity of PbD for data cleaning,
whereas more recent uses range from augmented reality prototyping \cite{leiva-2021-rapido}
and web automation \cite{chen-2023-miwa} to end-user software customizations
\cite{litt-2020-customization,litt-2020-wildcard}. Nevertheless, expressing conditions in PbD remains an
active research topic~\cite{pu-2022-semanticon,radensky-2018-conditionals}. A more general class of demonstrational
interfaces \cite{myers-2000-demonstrational} also includes programming by example, used
for example to infer data transformations in spreadsheets \cite{gulwani-2012-spreadsheets}.
Demonstrational systems can be used to directly perform actions, but also to generate code
that captures the interaction, as done e.g. in WREX \cite{drossos-2020-wrex}.

\subsubsection*{Incremental Recomputation}
Interactive programming systems with live previews \cite{petricek-2020-live,mcdirmid-2013-usable}
have been attempting to update the previews without full recomputation at least since the pioneering
work on the Cornell Program Synthesizer \cite{teitelbaum-1981-cps}. Outside live programming,
more work has been focused on updating computation when data change, although this can also
be source code of a program passed to an adaptive interpreter \cite{acar-2006-adaptive}.
Incremental recomputation is also a concern in notebook systems where cells can be run out of
order \cite{singer-2020-jollity}. This can be addressed by maintaining a dependency graph
\cite{petricek-2018-wrattler,koop-2017-dataflow} that also allows incremental recomputation
on code change.

\subsubsection*{Schema Change Control}
Schema change control is concerned with adapting data and code to reflect changes in schema.
The problem is well-studied in the context of databases \cite{brahmia-2024-evolution},
although only few systems also automatically adapt database queries \cite{wang-2019-schema}.
However, the problem is starting to be recognized in programming systems research
\cite{litt-2020-cambria,edwards-2025-schema} as well as live programming \cite{barenz-2020-live}
where state needs to be preserved during program editing.

@@

\subsubsection*{Concrete Programming}
Programming can be simplified by working with concrete values instead of abstractions, an idea
pioneered by the prototype-based programming language Self \cite{ungar-1987-self}. Self uses
prototypes at the object level. At the expression level, similar functionality can be provided
by managed copy \& paste. Subtext \cite{edwards-2006-copypaste,edwards-2022-copypaste} treats this
mechanism as central, whereas other systems view tracking of copy \& paste as an additional
editor featrue \cite{jablonski-2007-cren,toomim-2004-linked}. Notably, copy \& paste has been
also tracked in spreadsheets \cite{hermans-2015-copypaste} and Gridlets \cite{joharizadeh-2020-gridlets}
offer an appealing alternative design.

\subsubsection*{Other Programming Experiences}
Several experiences not directly addressed in this paper are worth further investigation.
Projectional editors such as Lorgnette \cite{gobert-2023-lorgnette},
live literals \cite{omar-2021-livelits}, projection boxes \cite{lerner-2020-boxes} and data detectors
\cite{nardi-1998-agents} allow visualization or editing of aspects of programs through a user interface.
The problem of making live and rich editors compositional is addressed by
Engraft \cite{horowitz-2023-engraft}. Programming environments may also integrate
live examples \cite{rauch-2019-babylonian} and a variety of AI assistantce tools \cite{petricek-2023-aias,blinn-2024-llms,mcnutt-2023-nbai}.

@@

In this section, we use the web-based prototype Webnicek to illustrate programming
experiences enabled by the Denicek substrate. Webnicek is based on a structure
editor that supports navigation in the document and issuing of edit commands. In this section, we follow two
formative examples (\S\ref{app:examples}) related to conference planning.

@@

\keyideabox{\faLightbulbO}{Programming by Demonstration}{Denicek implements programming by
demonstration (\S\ref{sec:impl-pbd}) by enabling the user to save and replay past interactions.
Those can be replayed directly, or used as event handlers in order to construct interactive user
interfaces (\S\ref{sec:impl-interaction}).}

\subsubsection*{\circled{D} Refactoring Document Structure} Another user starts with the initial
version of the document (A) and turns the list into a table. This is done by invoking a series of
commands that change tags, wrap elements, copy and transform values.

\subsubsection*{\circled{E} Merging Edits} The two document versions (C and D) can be automatically merged.
The refactoring is applied to all existing speakers. New speakers added using the ``Add speaker!''
button are also automatically added in the new format.

@@

\subsubsection*{\circled{F} Adding Budget Calculation} A third user adds a budget calculation to
the initial document. Formulas are represented as special {\small\Verb_x-formula_} nodes whose arguments
are other nodes or references to other nodes or formulas (highlighted on mouse hover).

\keyideabox{\faLightbulbO}{Incremental Recomputation}{Formulas are document nodes and
evaluating a formula yields additional edits that augment the document with the result
(\S\ref{sec:impl-eval}). Those evaluated edits are kept at the top of the document history and are
removed in case of a conflict (\S\ref{sec:impl-incremental}), resulting in the incremental
recomputation behavior.}

\subsubsection*{\circled{G} Merging Formulas} When the budget calculation is merged with the
other edits, references in formulas are automatically updated to point to the new list.
Adding new speaker via the ``Add speaker!'' button invalidates the evaluated result.

@@

\subsection{Selectors, Documents and Edits}
\label{sec:system-defs}
A computational document is a tree, consisting of four kinds of nodes (Fig.~\ref{fig:doc}).
Records and lists are labeled with a \emph{tag} as in HTML. Record fields have unique names.
Lists are expected to contain elements of the same structure, identified by a unique index (serving
as an ID). Primitive nodes can be strings, numbers or references to another location in the
document tree. References can be relative or absolute. Edit operations only use absolute
references (to denote a target node), but relative references can appear in the document (e.g., to
refer to another node within a list element).

A reference is represented as a sequence of selectors (Fig.~\ref{fig:doc}).
The document model assumes that lists are homogeneous and records heterogeneous, and so the
\ident{Any} selector makes it possible to refer to all elements of a list, but there is no
way to refer to all fields of a record. Note that Denicek does not use implicit numerical
indices for lists. The index of a new list item has to be supplied explicitly. The reasoning behind
this design choice is discussed in \S\ref{sec:discuss}.

\subsubsection*{Document Edits}
The supported document edits and their behavior are listed in Fig.~\ref{fig:edits}. All edits
require a \emph{target} to which they are applied. Targets are absolute references not containing the
\ident{Parent} selector. They can contain the \ident{Any} selector, in which case the edit is applied
to multiple nodes simultaneously. Most edits can only be applied to target node(s) of a specified
kind. As discussed in \S\ref{sec:discuss}, list elements as well as fields of a record are ordered
and edits that add a new item take the index or field name of a previous node.

The edits allow an arbitrary transformation of a document via a series of steps whose effect can be
tracked by the substrate. As we saw, Denicek updates references when document structure changes.
Fig.~\ref{fig:edits} distinguishes between edits that keep references in a document unchanged
(above) and edits that affect references (below).

@@

\subsubsection*{Automatic Reference Update}
As discussed in \S\ref{sec:discuss}, the structure of Denicek document is implicit and not
statically enforced (the substrate can be seen as dynamically typed). As a consequence, edits
that transform the structure can be used as structural edits (affecting the structure)
or as value edits (affecting the value of specific nodes). The latter can be used, for example,
when constructing an additional list item. An item can be first constructed and then populated
with values. This temporarily violates the invariant that collections are homogeneous
(an issue we might address by using transactions as discussed in \S\ref{sec:discuss}).
Another use of value edits is when evaluating a formula, which involves wrapping the formula
node (\S\ref{sec:impl-eval}), but the operation should not affect references to the formula
itself. To support value edits, edits that normally transform references (Fig.~\ref{fig:edits} below)
have a \emph{reference behavior} option that can be set to disable automatic reference
updating. This annotation is required when the target contains the \emph{Index} selector,
which is necessarilly a value edit affecting a specific node.

@@

An important principle of the Denicek design is that the effect an edit has on references inside
the document does not depend on the current value of the document, only its type. This makes it possible to
define merging solely in terms of edits, without reference to current document state.
% (The effect \ident{Copy} has on the document structure depends on the existing document
% structure, but not on its value).
This design choice makes it impossible to encode computational logic directly in the edits
(e.g., through conditional edits). As we will see in \S\ref{sec:impl-interaction}, such logic
can be provided as an additional mechanism on top of the underlying Denicek substrate.

@@

\subsubsection*{Applying Edits}
When applying an edit, Denicek locates the target node and transforms it according to the edit. If
the edit affects references in the document (Fig~\ref{fig:edits}, below), Denicek updates
the relevant references according to the rules shown in Fig.~\ref{fig:updates}, provided
that the reference behavior of the edit is not set to disable this.
%(this is required when %the target reference contains the \ident{Index} selector).

>>>>

\begin{abstract}
User-centric programming research gave rise to a variety of compelling programming experiences,
including collaborative source code editing, programming by demonstration, incremental
recomputation, schema change control, end-user debugging and concrete programming.
Those experiences advance the state of the art of end-user programming, but they are hard to
implement on the basis of established programming languages and system.

We contribute Denicek, a computational substrate that simplifies the implementation of the
above programming experiences. Denicek represents a program as a series of edits
that construct and transform a document consisting of data and formulas. Denicek provides three
operations on edit histories: edit application, merging of histories and conflict resolution.
Many programming experiences can be easily implemented by composing these three operations.

@@

A computational substrate defines the structures with which programs are constructed, how
the progam state is represented and how the state evolves during execution \cite{jakubovic-2022-ladder}.
The choice of a substrate affects what programming experiences can be readily supported.
For example, object-oriented programming has been historically linked to graphical user
interfaces~\cite{kay-1993-smalltalk}, while representing programs as lists enabled Lisp to become
a language laboratory \cite{steele-1993-lisp}.

@@

In this paper, we describe a programming substrate that makes it easy to develop diverse
programming experiences \cite{myers-2006-eup}. The experiences we focus on include:

\begin{itemize}
\item \emph{Collaborative editing.} Users should be able to locally modify a shared document
  and merge their changes, preferrably in a way that does not require a live connection to a central server~\cite{kleppmann-2019-local}.
\item \emph{Programming by Demonstration.} Allow users to construct simple programs by
  enacting the steps of the desired behavior using concrete examples and generalizing from the example~\cite{leiva-2021-rapido,cypher-1993-pbd}.
\item \emph{Incremental Recomputation.} When a part of a document changes, formulas whose result
  depends on the part are invalidated and, possibly automatically, recomputed \cite{teitelbaum-1981-cps,mcdirmid-2013-usable,horowitz-2023-engraft}.
\item \emph{Schema Change Control.} When the user evolves the structure of the document, affected
  data and formulas should automatically co-evolve to match the new structure~\cite{litt-2020-cambria,edwards-2025-schema}.
\item \emph{End-User Debugging.} The user should be able to ask provenance questions~\cite{cheney-2009-provenance} to understand
  why a computation resulted in a particular value and what inputs contributed to the result \cite{ko-2009-whyline}.
\item \emph{Concrete Programming.} It should be possible to reuse parts of program logic, or formulas,
  without introducing abstractions, that is, program against concrete values \cite{edwards-2006-copypaste,edwards-2022-copypaste}.
\end{itemize}

\subsubsection*{Two-Phase Methodology}
The technical focus of this paper fits within the interior mode of design science research \cite{adam-2021-dsr}.
To \emph{design} Denicek, we identify six \emph{formative examples} -- simple programming
tasks that manifest one or more of the desired programming experiences~(\S\ref{app:examples}).
Using those examples, we co-design the Denicek substrate and a simple web-based end-user programming
environment Webnicek~(\S\ref{sec:walk}), which is built directly on top of the substrate. Although
Webnicek can be used to complete end-user programming tasks, it is optimized for developing the
underlying substrate rather than for usability.

To \emph{evaluate} the usability of Denicek for the development of end-user programming systems,
we use it to build Datnicek (\S\ref{sec:case}), an innovative interactive data science notebook
that brings together a range of recent research work~\cite{kandel-2011-wrangler,drossos-2020-wrex,petricek-2022-thegamma,adams-2025-grove}.
We reflect on the development experience~(\S\ref{sec:case-reflection}) and provide a heuristic
evaluation of Denicek characteristics including importance and generality (\S\ref{sec:eval}).

@@

The key insight that we present in this paper is that many compelling programming experiences can
be implemented by leveraging this uniform document representation alongside with a suitable
composition of the three primitive operations.

Editing of data and formulas is done using a single set of primitive edit operations that
manipulate the document. A user-interface may provide a specialized editor, but still
trigger the primitive edits behind the scenes. Interacting with elements in the document,
such as a entering text in a textbox can also generate a document edit that can be merged or checked
for conflicts (\S\ref{sec:impl-interaction}).

As we will see, past edits that demonstrate an operation done with the
document can be recorded, allowing programming by demonstration (\S\ref{sec:impl-pbd}).  Replaying
such recorded edits is implemented using the merging operation, which means that recorded operations
continue working even if the document structure later evolves. Moreover, structural changes to the
document can be merged with concurrent data edits  (\S\ref{sec:impl-collab}). Evaluation of formulas
also generates document edits (\S\ref{sec:impl-eval}). If the evaluated edits conflict with manual
edits done later by the user, the evaluated edits are removed, yielding an incremental
recomputation behavior (\S\ref{sec:impl-incremental}).

@@

\begin{itemize}
\item We present the Denicek substrate (\S\ref{sec:system}) and provide a detailed description of its
  document representation, edit operations and the key three operations for working with edit histories.
\item We illustrate a range of end-user programming experiences supported in Webnicek, a simple
  web-based prototype programming system (\S\ref{sec:walk}), and discuss how the experiences are
  implemented on top of the Denicek substrate (\S\ref{sec:impl}).
\item We document important design decisions, alternatives and limitations (\S\ref{sec:discuss}).
  The analysis shows that the desired functionality requires a careful choice among interconnected
  design options.
\item To evaluate how Denicek simplifies the development of program\-ming systems, we
  build an innovative data science notebook Datnicek and assess its impementation
  complexity~(\S\ref{sec:case}). We also present heuristic evaluation of the system (\S\ref{sec:eval}).
\end{itemize}

\noindent
To enable others build on top of Denicek, we share our compact open-source
implementation at: \url{https://github/removed/for/review} %d3sprog/denicek

@@

\section{Background}
\label{sec:background}
Denicek aims to support a class of systems associated with end-user programming \cite{myers-2006-eup},
notational freedom and self-sustainability~\cite{jakubovic-2023-techdims}, liveness and interactivity
\cite{horowitz-2023-qualities,rein-2019-live}. We see programming more as interacting with a medium
or a substrate \cite{kay-1977-media,klokmose-2015-webstrates} than writing text \cite{rpg-2012-revolution}.
We focus on systems that allow gradual progression from a user to a developer, as envisioned
in Smalltalk~\cite{reenskaug-1981-byte} and use the term end-user programming loosely to refer to a
part of this spectrum, also including spreadsheet systems and notebooks for data science.
% We first review related systems and their technical design (\S\ref{sec:background-sys})
% before discussing programming experiences (\S\ref{sec:background-exp}).

@@

\subsubsection*{Programming Systems and Substrates}
A number of systems illustrate the qualities Denicek aims to support and have a related internal
structure. Subtext, BootstrapLab and Infra~\cite{edwards-2005-subtext,jakubovic-2022-ladder,hall-2017-infra}
use structured document-based program representation and provide some of the desired programming
experiences on top of this representation. Many of those design ideas can be traced back to
Boxer~\cite{disessa-1986-boxer}, which introduced the \emph{naive realism} principle (what the user
sees is all there is) that we also follow in Webnicek.

To indicate that Denicek is intended as an underlying infrastructure on top of which programming
systems can be built, we use the term \emph{computational substrate}. The term also dates back to
Boxer~\cite{disessa-1995-epistemology} and is related to the notion of dynamic media of Kay and
Goldberg \cite{kay-1977-media}. Webstrates \cite{klokmose-2015-webstrates}
revisit the idea, providing a substrate based on synchronization of documents (without edit
histories) that has been used as the basis for multiple programming systems \cite{radle-2017-codestrates,borowski-2022-varv}.

\subsubsection*{Edit Histories and Merging}
Manipulating programs through semantically meaningful edits is a technique used by structure
editors \cite{teitelbaum-1981-cps,hempel-2018-deuce,beckman-2023-sandblocks}. The langauge of
edits has been captured formally as an \emph{edit calculus} \cite{omar-2017-hazelnut} and
edit histories have also been recognized as a suitable basis for live programming environments \cite{storm-2013-deltas}.

Merging of edits is most frequently done in version control systems such as git. Source control ystems
based on a more rigorous design such as Pijul \cite{meunier-2024-pijul} delay merging to a later point
by using a graph or a lattice \cite{schurmann-2022-merging}. In the context of
programming environments, Grove \cite{adams-2025-grove} uses a commutative patches, with a graph
structure similar to Pijul, as the basis for a collaborative structure editor.

More generally, merging of edits can use the operational transform (OT) approach
\cite{davis-2002-otsgml}, where edit conflicts are reconciled, or conflict-free
representation (CRDTs) \cite{shapiro-2011-crdts,litt-2022-peritext}. The latter is commonly associated
with local-first software \cite{kleppmann-2019-local} that operates without a central server.
In both approaches, supporting complex edits on tree structures
remains a challenge \cite{jungnickel-2016-otjson,da-2024-jsoncrdt}. Mergeable replicated data
types (MRDTs) \cite{kaki-2019-mrdts} merge updates using a suitable relational
representation. Recent work on MySubstrates and Grove \cite{adams-2025-grove,klokmose-2024-mywebstrates}
has been based on CRDTs, whereas Edwards et al. \cite{edwards-2021-typed} prefer OT.

@@

\subsubsection*{Collaborative Editing}
Since the early collaborative programming environments such as Collabode \cite{goldman-2011-collaborative},
real-time collaboration has became widely used, if not always without challenges \cite{tan-2024-vslive}.
Merging concurrent edits is one such challenge. In addition to OT-based and CRDT-based
approaches \cite{klokmose-2024-mywebstrates}, conflicts arising during
collaboration have also been solved using fine-grained locking \cite{wang-2024-nbconflicts}.

\subsubsection*{Programming by Demonstration}
Earliest PbD systems used the paradigm for tasks ranging from graphics
and user interfaces to general-purpose programming \cite{smith-1975-pygmalion,cypher-1993-pbd}.
Wrangler \cite{kandel-2011-wrangler} showed the effectivity of PbD for data cleaning,
whereas more recent uses range from augmented reality prototyping \cite{leiva-2021-rapido}
and web automation \cite{chen-2023-miwa} to end-user software customization
\cite{litt-2020-customization,litt-2020-wildcard}. Expressing conditions in PbD remains an
active research topic~\cite{pu-2022-semanticon,radensky-2018-conditionals}. A more general class of demonstrational
interfaces \cite{myers-2000-intelligence} also includes programming by example, used
for example for data transformation in spreadsheets \cite{gulwani-2012-spreadsheets}.
Demonstrational interfaces can be used to directly perform actions, but also to generate code
as in Wrex \cite{drossos-2020-wrex}.

\subsubsection*{Incremental Recomputation}
Interactive programming systems with live previews \cite{petricek-2020-live,mcdirmid-2013-usable}
have been attempting to update the previews without full recomputation at least since the pioneering
work on the Cornell Program Synthesizer \cite{teitelbaum-1981-cps}. Outside live programming,
more work has been focused on updating computation when data change, although such data can also
be source code passed to an adaptive interpreter \cite{acar-2006-adaptive}.
Incremental recomputation is also a concern in notebook systems where cells can be run out of
order \cite{singer-2020-jollity}. The ordering problem can be addressed using a dependency graph
\cite{petricek-2018-wrattler,koop-2017-dataflow}, allowing for incremental recomputation
on code change.

\subsubsection*{Schema Change Control}
Schema change control is concerned with adapting data and code to reflect changes in schema.
The problem is well-studied in the context of databases \cite{brahmia-2024-evolution},
although only few systems also automatically adapt database queries \cite{wang-2019-schema}.
The problem is starting to be recognized in programming systems research
\cite{litt-2020-cambria,edwards-2025-schema} as well as live programming \cite{barenz-2020-live}
where state needs to be preserved during program editing.

@@

\subsubsection*{Concrete Programming}
Programming can be simplified by working with concrete values instead of abstractions, an idea
pioneered by the prototype-based programming language Self \cite{ungar-1987-self}. In Self,
prototypes are used at the object level. At the expression level, similar functionality can be provided
by managed copy \& paste. Subtext \cite{edwards-2006-copypaste,edwards-2022-copypaste} treats this
mechanism as central, whereas other systems view tracking of copy \& paste as an extra
editor featrue \cite{jablonski-2007-cren,toomim-2004-linked}. Copy \& paste has also been
tracked in spreadsheets \cite{hermans-2015-copypaste}. Gridlets \cite{joharizadeh-2020-gridlets}
offer a spreadsheet abstraction based on reusing concrete computation, similar to the one proposed
for The Gamma~\cite{petricek-2020-live}.

\subsubsection*{Other Programming Experiences}
Several experiences not directly addressed in this paper are worth further investigation.
Projectional editors such as Lorgnette \cite{gobert-2023-lorgnette},
live literals \cite{omar-2021-livelits}, projection boxes \cite{lerner-2020-boxes} and data detectors
\cite{nardi-1998-agents} allow visualization or editing of aspects of programs through a user interface.
The problem of making live and rich editors compositional is addressed by
Engraft \cite{horowitz-2023-engraft}. Programming environments may be also improved by integrating
live examples \cite{rauch-2019-babylonian} and a variety of AI assistantce
tools \cite{petricek-2023-aias,blinn-2024-llms,mcnutt-2023-nbai}.

@@

In this section, we use the web-based prototype Webnicek to illustrate programming
experiences enabled by the Denicek substrate. Webnicek is based on a structure
editor that supports navigation in the document and issuing of edit commands. We follow two
formative examples (\S\ref{app:examples}) where users collaboratively plan a conference.

@@

\keyideabox{\faLightbulbO}{Programming by Demonstration}{Denicek implements programming by
demonstration (\S\ref{sec:impl-pbd}) by enabling the user to save and replay past interactions.
Past interactions can be replayed directly, or used as event handlers in order to construct interactive user
interfaces (\S\ref{sec:impl-interaction}).}

\subsubsection*{\circled{D} Refactoring Document Structure} Another user starts with the initial
version of the document (A) and turns the list into a table. This is done by invoking a series of
commands that change tags, wrap elements, copy values and transform them.

\subsubsection*{\circled{E} Merging Edits} The two document versions (C and D) can be automatically merged.
The refactoring is applied to all existing speakers. New speakers added using the ``Add speaker!''
button are also automatically transformed to the new format.

@@

\subsubsection*{\circled{F} Adding Budget Calculation} A third user adds a budget calculation to
the initial document. Formulas are represented as nodes with the {\small\Verb_x-formula_} tag,
whose arguments are other nodes, references to other nodes, or formulas (highlighted on mouse hover).

\keyideabox{\faLightbulbO}{Incremental Recomputation}{Formulas are document nodes.
Evaluating a formula yields edits that augment the document with the result
(\S\ref{sec:impl-eval}). Those evaluated edits are kept at the top of the document history and are
removed in case of a conflict (\S\ref{sec:impl-incremental}), a logic that implements
incremental recomputation using core Denicek operations.}

\subsubsection*{\circled{G} Merging Formulas} When the budget calculation (F) is merged with the
earlier edits (E), references in formulas are automatically updated to point to the new list.
Adding new speaker via the ``Add speaker!'' button invalidates the evaluated result.

@@

\subsection{Selectors, Documents and Edits}
\label{sec:system-defs}
A computational document is a tree, consisting of four kinds of nodes (Fig.~\ref{fig:doc}).
Records and lists are labeled with a \emph{tag} as in HTML. Record fields have unique names.
Lists are expected to contain elements of the same structure, identified by a unique index (serving
as an ID). Primitive nodes can be strings, numbers or references to another location in the
document tree. References can be relative or absolute. Edit operations only use absolute
references (to denote a target node), but relative references can appear in the document (e.g., to
refer to another node within the same list element).

A reference is represented as a sequence of selectors (Fig.~\ref{fig:doc}).
The document model assumes that lists are homogeneous and records heterogeneous, and so the
\ident{Any} selector makes it possible to refer to all elements of a list, but there is no
way to refer to all fields of a record. Because Denicek does not use implicit numerical
indices for lists, the index of a new list item has to be supplied explicitly. The reasoning behind
this design choice is discussed in \S\ref{sec:discuss}.

\subsubsection*{Document Edits}
The supported document edits and their behaviors are listed in Fig.~\ref{fig:edits}. All edits
require a \emph{target} to which they are applied. Targets are absolute references not containing the
\ident{Parent} selector. They can contain the \ident{Any} selector, in which case the edit is applied
to multiple nodes simultaneously. Most edits can only be applied to target node(s) of a specified
kind. As discussed in \S\ref{sec:discuss}, list elements as well as fields of a record are ordered
and edits that add a new item take the index or field name of a previous node.

The edits can transform the document structure. Denicek tracks the effect of the edits on the
structure and updates references when the document structure changes.
Fig.~\ref{fig:edits} distinguishes between edits that keep references in a document unchanged
(above) and edits that can affect references (below).

@@

\subsubsection*{Automatic Reference Update}
As discussed in \S\ref{sec:discuss}, the structure of Denicek document is implicit and not
statically enforced (the substrate can be seen as being dynamically typed). As a consequence, edits
that transform the structure can serve both as structural edits (affecting the structure)
and as value edits (affecting the value of specific nodes). The latter can be used, for example,
when constructing an additional list item. An item is first constructed and then populated
with values. This temporarily violates the invariant that collections are homogeneous
(an issue that could be address by using transactions as discussed in \S\ref{sec:discuss}).

Another use of value edits is when evaluating a formula, which involves wrapping the formula
node (\S\ref{sec:impl-eval}), an operation that should not affect references to the formula
itself. To support value edits, edits that normally transform references (Fig.~\ref{fig:edits} below)
have a \emph{reference behavior} option that can disable automatic reference
updating. This annotation is required when the target contains the \emph{Index} selector,
which is necessarilly a value edit affecting a specific node.

@@

An important principle of the Denicek design is that the effect an edit has on references inside
the document does not depend on the current value of the document. This makes it possible to
define merging solely in terms of edits, without reference to current document state.
% (The effect \ident{Copy} has on the document structure depends on the existing document
% structure, but not on its value).
This design choice makes it impossible to encode computational logic directly in the edits
(e.g., through conditional edits). As we will see in \S\ref{sec:impl-interaction}, such logic
can be provided as an additional mechanism on top of the underlying Denicek substrate.

@@

\subsubsection*{Applying Edits}
When applying an edit, Denicek locates the target node and transforms it according to the edit. If
the edit affects references in the document (Fig~\ref{fig:edits}, below), Denicek updates
the relevant references according to the rules shown in Fig.~\ref{fig:updates}, provided
that the reference behavior of the edit does not prevent updating.
%(this is required when %the target reference contains the \ident{Index} selector).
